Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-12 draft.tex
Revised Version of Bernhard

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the aging society.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by an radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

@@

Closest to ours is the work of Gla{\ss}er \cite{Glasser2013} and Niemann et al. \cite{Niemann2014}, who both use decision rules for their analysis.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

@@

%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$. We refer to total curvature from now on as \emph{curvature}.
%%
	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$. We refer to total torsion from now on as \emph{torsion}
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

@@

\paragraph{Attribute Mapping}
The scatter plot axis are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
%%
We can support this by color coding the data points according to group affiliation.

@@

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variables encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry on the visualization, which then displays the corresponding decision tree in detail.
%%
This allows for sequentially analyzing the classifications.

@@

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\subsection{Applicability}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

>>>>

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP} gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
Cohort studies are an instrument of epidemiological research.
%For epidemiological research, cohort studies are carried out.
%%
Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent diseases in the western civilization and gains importance due to the aging society.
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

A survey on image-centric cohort studies and strategies to analyze the resulting data is given in \cite{Preim2014}.

@@

Closest to ours is the work of Gla{\ss}er et al. and Niemann et al., who both use decision rules for their analysis \cite{Glasser2013,Niemann2014}.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
Each subjects' model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

@@

%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$. We refer to total curvature from now on as \emph{curvature}.
%%
	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$. We refer to total torsion from now on as \emph{torsion}.
%%
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

@@

\paragraph{Attribute Mapping}
%%
The scatter plot axis are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
%If we for example classify and compare the same variables for different subject groups, for example male and female subjects.
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
%Classification of the same variable 
%%
%We can support this by color coding the data points according to group affiliation.

@@

%\com{ToDo: SQRT Skalierung der Treesize-Dimension, da hauptsaechlich geringe Groessen von interesse sind, ausreisser aber diesen Bereich schwer durchschaubar machen}
% %%
%
% %%
% Abstracting the classification results into two continuos variables allows us to visualize the result of 134 classifications in one scatter plot.
% %%
% By mapping the group affiliation (in our case body-size) to color or shape, we can even plot all 402 results in a single visualization (Fig.~\ref{fig:decision-tree} a).
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several binary variables.
%%
Each binary variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four binary variables (\emph{No pain - Yes/No} to \emph{Large Pain - Yes/no}).
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data parameter.
%%\com{ToDo: Dummy-Variables}
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry on the visualization, which then displays the corresponding decision tree in detail.
%%
This allows for sequentially analyzing the classifications.

@@

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\subsection{Applicability}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%
